An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-09 paper_versions/nips_2016.tex
Cleaning up directory by removing nips files

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, M-H requires
calculations over the full dataset to produce a new sample.  Recent
results from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional uniform random variable
$u$. The amount of data consumed for each test varies significantly
from one minibatch to the next, and depends on the current sample, the
proposed sample, and on the random variable $u$. By
contrast,~\cite{conf/uai/MaclaurinA14}\cite{TallData15} perform exact
tests but require a lower bound on parameter distribution across its
domain.  The amount of data reduction depends on the accuracy of this
bound, and such bounds are only available for relatively simple
distributions. By contrast our approach is ``black box'' and requires
only statistics gathered from the minibatches as they are used.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test}. Because of this, the amount of
data required for each test is fixed with high probability and in expected
value.
%% {\color{blue} Daniel: we really need to show the amount of data is fixed
%% with ``high probability'' (the results suggest ``OK probability'').} 
%% Analysis show data amount is gaussian and short-tailed and therefore bounded with high p
%% experiments back this up. 
We use a Barker test function~\cite{Barker65} rather than a Metropolis
test, which makes our test naturally error tolerant. The idea of using
a noise-tolerant test using Barker's test function was suggested but
not explored empirically in~\cite{Bardenet15} section 6.3. But the
asymptotic test statistic CDF and the Barker function are different,
which leads to fixed errors for the approach
in~\cite{Bardenet15}. Here we show that the difference between the
distributions can be corrected with an additive random variable. This
leads to a test which is fast, and whose error can be made arbitrarily
small.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$,
we can write
\begin{equation}\label{eq:relationship}
    \Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is
approximately normal with variance $\sigma^2(\Delta^*)$.  But to perform the
test in Equation~(\ref{eq:the_exact_test}) we want $\Delta + X$ for a logistic
random variable $X$ (call it $X_{\rm log}$ from now on). In~\cite{Bardenet15} it
was proposed to use $\Delta^*$ in a Barker test anyway and tolerate the fixed
error caused by this approximation.

@@

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative
by inspection of equation (\ref{eq:whyneg}) below.
This implies that we can find a $u \in (0,1)$ so that
$\psi(u,\theta,\theta') = \log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.
Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$ and Equation~(\ref{eq:lemma_ll_ratio}), we see that
\begin{equation}
  \log u_0 = N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right]
\end{equation}
\begin{equation}\label{eq:whyneg}
    log u_0 = -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{equation}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$ used in ~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} , where $\not\approx$
means ``significantly different from'' under the distribution over
samples of $x_i$. This turns out to be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0. \label{eq:accept_test_zero}
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.

>>>>

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, M-H requires
calculations over the full dataset to produce a new sample.  Recent
results from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional uniform random variable
$u$. The amount of data consumed for each test varies significantly
from one minibatch to the next, and depends on the current sample, the
proposed sample, and on the random variable $u$. By
contrast,~\cite{conf/uai/MaclaurinA14,TallData15} perform exact
tests but require a lower bound on parameter distribution across its
domain.  The amount of data reduction depends on the accuracy of this
bound, and such bounds are only available for relatively simple
distributions. By contrast our approach is ``black box'' and requires
only statistics gathered from the minibatches as they are used.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test}. Because of this, the amount of
data required for each test is fixed with high probability and in expected
value.
%% {\color{blue} Daniel: we really need to show the amount of data is fixed
%% with ``high probability'' (the results suggest ``OK probability'').} 
%% Analysis show data amount is gaussian and short-tailed and therefore bounded with high p
%% experiments back this up. 
We use a Barker test function~\cite{Barker65} rather than a Metropolis
test, which makes our test naturally error tolerant. The idea of using
a noise-tolerant test using Barker's test function was suggested but
not explored empirically in~\cite{TallData15} section 6.3. But the
asymptotic test statistic CDF and the Barker function are different,
which leads to fixed errors for the approach
in~\cite{TallData15}. Here we show that the difference between the
distributions can be corrected with an additive random variable. This
leads to a test which is fast, and whose error can be made arbitrarily
small.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$,
we can write
\begin{equation}\label{eq:relationship}
    \Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is
approximately normal with variance $\sigma^2(\Delta^*)$.  But to perform the
test in Equation~(\ref{eq:the_exact_test}) we want $\Delta + X$ for a logistic
random variable $X$ (call it $X_{\rm log}$ from now on). In~\cite{TallData15} it
was proposed to use $\Delta^*$ in a Barker test anyway and tolerate the fixed
error caused by this approximation.

@@

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative
by inspection of equation (\ref{eq:whyneg}) below.
This implies that we can find a $u \in (0,1)$ so that
$\psi(u,\theta,\theta') = \log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.
Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$ and Equation~(\ref{eq:lemma_ll_ratio}), we see that
\begin{equation}
  \log u_0 = N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right]
\end{equation}
\begin{equation}\label{eq:whyneg}
    \log u_0 = -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{equation}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$ used in ~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} , where $\not\approx$
means ``significantly different from'' under the distribution over
samples of $x_i$. This turns out to be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0. \label{eq:accept_test_zero}
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.
